---
title: "R for data science"
author: "JORGE GARCIA"
date: '2022-05-22'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE) # turn off warnings and messages

options(scipen=999) # turn off scientific notation
```


```{r}
# load data viz/analysis packages 
library(tidyverse)
library(janitor)

# load math operations packages
library(PlaneGeometry)

# load animation paclages
library(gganimate)  # to create animations
library(gifski)     # to render animations
library(png)        # to save animations

```

## Import data
readr supports the following file formats with these read_*() functions:

- comma-separated values (CSV) files
- tab-separated values (TSV) files
- delimited files (CSV and TSV are important special cases)
- fixed-width files
- whitespace-separated files
- web log files

```{r}
censos_bolivia <- read_csv("data-raw/censos_bolivia.csv")

print(censos_bolivia)
```
## Tidy data - Variable Names
`rename()` function changes the names of individual variables using `new_name = old_name` syntax. 

Best practices: 

- Follow conventions (lower case, camel format, etc.).
- Avoid blank spaces.
- In English. (?)
- Short.

```{r}
censos_bolivia_tidy <- censos_bolivia %>%
  rename(year = Año,
         pob_total = `Población Total`,
         pob_urban = `Población urbana`,
         pob_rural = `Población rural`)

```

## Tidy data - Wide and Long Data
`pivot_longer()` "lengthens" data, increasing the number of rows and decreases the number of columns. The inverse transformation is `pivot_wider()`

Best practices:

- Reduce number of variables (to perform summaries with less code).
- Cleaning up data is often going from wide to long data. 
- Each column should represent a variable, and each row should represent an observation an observation. 

```{r}
censos_bolivia_tidy <- censos_bolivia %>%
  rename(year = Año,
         pob_total = `Población Total`,
         pob_urban = `Población urbana`,
         pob_rural = `Población rural`) %>% 
  pivot_longer(cols = -year,
               names_to = "pob_type", # population type
               values_to = "value")

print(censos_bolivia_tidy)
```
## Tidy data - Recode Variables 
`mutate()` adds new variables and preserves existing ones. 

```{r}
censos_bolivia_tidy <- censos_bolivia %>%
  rename(year = Año,
         pob_total = `Población Total`,
         pob_urban = `Población urbana`,
         pob_rural = `Población rural`) %>% 
  pivot_longer(cols = -year,
               names_to = "pob_type",
               values_to = "value") %>%
  mutate(pob_type = case_when(
    pob_type == "pob_total" ~ "total",
    pob_type == "pob_urban" ~ "urban",
    pob_type == "pob_rural" ~ "rural"
 ))

```

## Plotting Data
`ggplot()` initializes a ggplot object. It can be used to declare the input data frame for a graphic and to specify the set of plot aesthetics intended to be common throughout all subsequent layers unless specifically overridden

```{r}
censos_bolivia_tidy %>% 
  ggplot(aes(
    x = year, 
    y = value,
    group = pob_type,
    )) +
  geom_area() # geom_point, geom_area
```

### Customizing Plots - Color

```{r}
censos_bolivia_tidy %>% 
  ggplot(aes(
    x = year, 
    y = value,
    group = pob_type,
    )) +
  geom_line(aes(
    color = pob_type
    ))
```

### Customizing Plots - Color HEX Values

`scale_manual` These functions allow you to specify your own set of mappings from levels in the data to aesthetic values.

- Also accepts HTML HEX value as input

```{r}
censos_bolivia_tidy %>% 
  ggplot(aes(
    x = year, 
    y = value,
    group = pob_type,
    )) +
  geom_line(aes(color = pob_type)) +
  scale_color_manual(values = c("Green", "Yellow", "Blue"))


censos_bolivia_tidy %>% 
  ggplot(aes(
    x = year, 
    y = value,
    group = pob_type,
    color = pob_type
    )) +
  geom_line() +
  scale_color_manual(values = c("#cc0000", "#3cb371", "#71a6d2"))
```

### Customizing Plots - Remove Default Gray Brackground and Customize 

`theme_()` can be used if you just need to tweak the display of an existing theme. Themes are a powerful way to customize the non-data components of your plots.

```{r}
censos_bolivia_tidy %>% 
  ggplot(aes(
    x = year, 
    y = value,
    group = pob_type,
    color = pob_type
    )) +
  geom_line(size = 0.4) +
  scale_color_manual(values = c("#cc0000", "#3cb371", "#71a6d2")) + 
  theme_minimal()+
  geom_point()
```

### Customizing plots - Axis Titles

```{r}
censos_bolivia_tidy %>% 
  ggplot(aes(
    x = year, 
    y = value,
    group = pob_type,
    color = pob_type
    )) +
  geom_line(size = 0.4) +
  scale_color_manual(values = c("#cc0000", "#3cb371", "#71a6d2")) + 
  theme_minimal() +
  geom_point() + 
  scale_x_continuous(name = "Year") +
  scale_y_continuous(labels = scales::unit_format(unit = "M", scale = 1e-6))
```

### Customizing plots - Adding Labels
Best practices:

- Good labels are critical for making your plots accessible to a wider audience. 
- Ensure the axis and legend labels display the full variable name.
- `labs()` function.

```{r}
censos_bolivia_tidy %>% 
  ggplot(aes(
    x = year, 
    y = value,
    group = pob_type,
    color = pob_type
    )) +
  geom_line(size = 0.4) +
  scale_color_manual(values = c("#cc0000", "#3cb371", "#71a6d2")) + 
  theme_minimal() +
  geom_point() + 
  scale_x_continuous(breaks = c(1976,1992,2001,2012)) +
  scale_y_continuous(labels = scales::unit_format(unit = "M"
                                                  , scale = 1e-6)) + 
  labs(title = "Population by area of residence in Bolivia from 1976 to 2012",
       subtitle = "Urban and rural population",
       y = "Population (millions)",
       x = "Year",
       caption = "Sources: INE")
```

## Finding and Adding Insights 

- `annotate()`
- Get insights from data available 
- Year where Rural and Urban population intersect.
- create annotation layers with insights, passed in as vectors. 
- Avoid adding columns or rows into the data frame. 
- Accepts control characters `\r\n` `\r` and `\n`

### Intersection of Rural and Urban Populations 

Finding year when urban population grows faster than rural

```{r}
# finding when urban population grows faster than rural
lineurban <- Line$new(A = c(1976, 1906324), B = c(1992, 3694846))
linerural <- Line$new(A = c(1976, 2707095), B = c(1992, 2725946))

intersectionLineLine(lineurban, linerural)
```


```{r}
censos_bolivia_tidy %>% 
  ggplot(aes(
    x = year, 
    y = value,
    group = pob_type,
    color = pob_type
    )) +
  geom_line(size = 0.4) +
  scale_color_manual(values = c("#cc0000", "#3cb371", "#71a6d2")) + 
  theme_minimal() +
  geom_point() + 
  scale_x_continuous(breaks = c(1976,1992,2001,2012)) +
  scale_y_continuous(labels = scales::unit_format(unit = "M"
                                                  , scale = 1e-6)) + 
  labs(title = "Population by area of residence in Bolivia",
       subtitle = "Urban and rural population",
       y = "Population (millions)",
       x = "Year",
       caption = "Sources: INE") +
  # urban vs rural population comparison
  annotate("text",
          x = 1983.24,
          y = 3600000,
          size = 3,
          label = "Urban population starts\nto grow faster than rural\nin 1983",
          color = "#778899") +
  # total population since 1976
  annotate("text",
           x = 1976,
           y = 5700000, 
           hjust = 0.2,
           size = 3,
           label = "Total population\nis 4.6 M in 1976",
           color = "#778899")
```

### Doubling of population with respect to 1976

Finding year where population doubles using total population in 1976 as reference. 

```{r}
# Finding when population doubles
 4613419 * 2 # Hallar el valor de duplicacion 

linetotal <- Line$new(A = c(1976, 4613419), B = c(2012, 10059856)) # Poblacion total
linedouble <- Line$new(A=c(1976, 9226838), B = c(2012, 9226838))  # Duplicacion 

intersectionLineLine(linetotal, linedouble)
```

```{r}
censos_bolivia_tidy %>% 
  ggplot(aes(
    x = year, 
    y = value,
    group = pob_type,
    color = pob_type
    )) +
  geom_line(size = 0.4) +
  scale_color_manual(values = c("#cc0000", "#3cb371", "#71a6d2")) + 
  theme_minimal() +
  geom_point() + 
  scale_x_continuous(breaks = c(1976,1992,2001,2012)) +
  scale_y_continuous(labels = scales::unit_format(unit = "M"
                                                  , scale = 1e-6)) + 
  labs(title = "Population by area of residence in Bolivia",
       subtitle = "Urban and rural population",
       y = "Population (millions)",
       x = "Year",
       caption = "Sources: INE") +
  # urban vs rural population comparison
  annotate("text",
          x = 1983.24,
          y = 3600000,
          #hjust = 0,
          size = 3,
          label = "Urban population starts\nto grow faster than rural\nin 1983",
          color = "#778899") +
  # total population since 1976
  annotate("text",
           x = 1976,
           y = 5700000, 
           hjust = 0.2,
           size = 3,
           label = "Total population\nis 4.6 M in 1976",
           color = "#778899") +
  # total population doubles in 2006
  annotate("text",
           x = 2006,
           y = 9900000,
           size = 3,
           color = "#778899",
           label = "Population doubles\nto 9.2 M in 2006.") 
```

## Animations
`gganimate()` A Grammar of Animated Graphics that helps create static visualizations compatible with ggplot, to specify transitions and animations. 

```{r}
censos_bolivia_tidy_gif <- censos_bolivia_tidy %>% 
  ggplot(aes(
    x = year, 
    y = value,
    group = pob_type,
    color = pob_type
  )) +
  geom_line(size = 0.4) +
  scale_color_manual(values = c("#cc0000", "#3cb371", "#71a6d2")) + 
  theme_minimal() +
  geom_point() + 
  scale_x_continuous(breaks = c(1976,1992,2001,2012)) +
  scale_y_continuous(labels = scales::unit_format(unit = "M"
                                                  , scale = 1e-6)) + 
  labs(title = "Population by area of residence in Bolivia",
       subtitle = "Urban and rural population",
       y = "Population (millions)",
       x = "Year",
       caption = "Sources: INE") +
  # urban vs rural population comparison
  annotate("text",
           x = 1983.24,
           y = 3600000,
           #hjust = 0,
           size = 3,
           label = "Urban population starts\nto grow faster than rural\nin 1983",
           color = "#778899") +
  # total population since 1976
  annotate("text",
           x = 1976,
           y = 5700000, 
           hjust = 0.2,
           size = 3,
           label = "Total population\nis 4.6 M in 1976",
           color = "#778899") +
  # total population doubles in 2006
  annotate("text",
           x = 2006,
           y = 9900000,
           size = 3,
           color = "#778899",
           label = "Population doubles\nto 9.2 M in 2006.") +
  transition_reveal(year) # <- line of code to create animations

censos_bolivia_tidy_gif
```
